---
title: "EPL Simulation"
author: "Jack Letcher, Treva Winlock, Andrew Brill"
date: "March 12, 2018"
output: html_document
---


# English Premier League 1998-2018
## Source: http://www.football-data.co.uk/englandm.php

The Poisson distribution applies when: (1) the event is something that can be counted in whole numbers; (2) occurrences are independent, so that one occurrence neither diminishes nor increases the chance of another; (3) the average frequency of occurrence for the time period in question is known; and (4) it is possible to count how many events have occurred, such as the number of times a team wins or loses during a given season. The Poisson Distribution gives the expected frequency profile for events. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, results="hide", include=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
```

## Temporary Home/Away Tables
```{r temp tables, results="hide"}
# Read in local .csv file 
EPL <- read.csv("EPL.csv")

# Format "Date" vector
EPL$Date <- as.Date(dmy(EPL$Date))
EPL$Date <- as.Date(EPL$Date, "%Y/%m/%d")

# Create Match and Team variables
match_table <- paste(EPL$HomeTeam, EPL$AwayTeam, sep = " - ")
team_table <- unique(c(EPL$HomeTeam, EPL$AwayTeam)) %>% sort

# Create home table (Games Played, Pts., Goals Scored, Goals Conceded)
home_table <- EPL %>% 
          group_by(HomeTeam) %>%
          summarise(P = length(FTR),
                    Pts = sum((FTHG > FTAG) * 3 + (FTHG == FTAG) * 1),
                    GS = sum(FTHG),
                    GC = sum(FTAG)) %>%
          ungroup()

# Used for a lookup table 
home_table <- cbind(home_table, Team_Num = 1:47)

# Create away table (Games Played, Pts., Goals Scored, Goals Conceded)
away_table <- EPL %>% 
          group_by(AwayTeam) %>%
          summarise(P = length(FTR),
                    Pts = sum((FTHG < FTAG) * 3 + (FTHG == FTAG) * 1),
                    GS = sum(FTAG),
                    GC = sum(FTHG)) %>%
          ungroup()

# Used for a lookup table
away_table <- cbind(away_table, Team_Num = 1:47)

# Create consolidated table with Total Goals Scored/Conceded
EPL.team.stats <- data.frame(Team = team_table,
                            Points = home_table$Pts + away_table$Pts,
                            GD = (home_table$GS + away_table$GS) - (home_table$GC + away_table$GC),
                            TGS = (home_table$GS + away_table$GS)/(home_table$P + away_table$P),
                            TGC = (home_table$GC + away_table$GC)/(home_table$P + away_table$P), stringsAsFactors = FALSE)
```

### Visualizations
#### Home and Away Team Plot Charts (Games Played, Points, Goals Scored, Goals Conceded)
```{r plot_viz, echo=FALSE}
# Visualize home and away tables
plot(home_table)
plot(away_table)
```

## Matches To Be Played
```{r TBP}
# Create new EPL table that normalizes Goals Scored and Goals Conceded
# Create Expected Home Goals and Expected Away Goals
EPL.new <- expand.grid(HomeTeam = team_table, AwayTeam = team_table, stringsAsFactors = FALSE) %>%
            filter(HomeTeam != AwayTeam) %>%
            mutate(Match = paste(HomeTeam, AwayTeam, sep = " - ")) %>%
            filter(!(Match %in% match_table)) %>%
            select(-Match) %>%
            mutate(HG = mean(EPL$FTHG),
                   AG = mean(EPL$FTAG),
                   TG = (mean(EPL$FTHG) + mean(EPL$FTAG))/2) %>%
            right_join(subset(EPL.team.stats, select = -c(Points, GD)),  by = c("HomeTeam" = "Team")) %>%
            right_join(subset(EPL.team.stats, select = -c(Points, GD)), by = c("AwayTeam" = "Team")) %>%
            setNames(c("HomeTeam", "AwayTeam", "HG", "AG", "TG", 
                       "GSBH", "GCBH", "GSBA", "GCBA")) %>%
            mutate(ExpHG = (GSBH / TG) * (GCBA / TG) * (HG / TG) * TG,
                   ExpAG = (GSBA / TG) * (GCBH / TG) * (AG / TG) * TG) %>%
            ungroup()
```

## Season Simulation
```{r simulation, results="hide"}
# Number of games simulated
iSimulation <- 500
n <- length(team_table)

# Create new shell for simulation data
EPL.all <- data.frame(Team = rep(team_table, iSimulation),
                     SimNo = rep(1:iSimulation, each = n),
                     Pts = rep(NA, n * iSimulation),
                     GD = rep(NA, n * iSimulation),
                     Rank = rep(NA, n * iSimulation))

# Progress Bar for simulation
win_bar <- winProgressBar(title = "Churning Through Your Simulation", 
                     label = "Simulating ... 0% done", 
                     min = 0, 
                     max = iSimulation, 
                     initial = 0)

# For loop that simulates through preset number of loops
for (i in 1:iSimulation){

# Random Generation for Poisson Distribution -- rpois
# Lambda representing vector of non-negative means         
  temp_table <- EPL.new %>% 
              mutate(x1 = rpois(nrow(EPL.new), lambda = EPL.new$ExpHG), 
                     x2 = rpois(nrow(EPL.new), lambda = EPL.new$ExpAG), 
                     HPts = 3 * (x1 > x2) + 1 * (x1 == x2),
                     APts = 3 * (x1 < x2) + 1 * (x1 == x2))
# Results table   
  results <- EPL.team.stats %>% select(Points, GD) + 
         temp_table %>% group_by(HomeTeam) %>% summarise(Pts = sum(HPts),
                                                  GD = sum(x1) - sum(x2)) %>% select(Pts, GD) + 
         temp_table %>% group_by(AwayTeam) %>% summarise(Pts = sum(APts),
                                                  GD = sum(x2) - sum(x1)) %>% select(Pts, GD) 

  EPL.all[(n*(i-1) + 1):(n*i), c("Pts", "GD")] <- results
  
  results$PGD <- results$Points + (results$GD - min(results$GD) + 1) / max((results$GD - min(results$GD) + 1) + 1)
  EPL.all[(n*(i-1) + 1):(n*i), c("Rank")] <- rank(-results$PGD, ties.method = "random")  

# Progress Bar for simulation results  
  info <- sprintf("Simulating ... %d%% complete", round((i/iSimulation)*100))
  setWinProgressBar(win_bar, i, label = info)  
}
close(win_bar)
```

 
```{r teamname_addition, echo=FALSE}
# Create new table to house lookup values for teamnames
new <- EPL.all
new[] <- home_table$HomeTeam[match(unlist(EPL.all), home_table$Team_Num)]

# Combine tables to clean up numeric teams
EPL.all.new <- cbind(Team_Name = new[1], EPL.all)
EPL.all.new <- EPL.all.new[-c(2)]
```

## Winner
```{r winner model}
# Output for Position, also provides probability
EPL.all.new %>% filter(Rank == 1) %>% select(Team) %>% table/iSimulation
```

## Arsenal Projected Finish
```{r projected }
# Output for Team, also provides probability
EPL.all.new %>% filter(Team == "Arsenal") %>% select(Rank) %>% table/iSimulation
```

## Arsenal Point Distribution
```{r point distribution}
# Output for Point Distribution, also provides probability
EPL.all.new %>% filter(Team == "Arsenal") %>% select(Pts) %>% table/iSimulation
```
## Table of Probabilities per Position (Top 9)
```{r Probabilities}
# Full table results by Team and probabilties of finish place
full_team_table <- table(EPL.all.new$Team, EPL.all.new$Rank)/iSimulation
full_team_table <- full_team_table[,1:9]
full_team_table
```

```{r addteam, echo=FALSE}
# Add Team Names to stats table
Team_Names <- data.frame(home_table$HomeTeam)
EPL.team.stats <- cbind(Team_Names, EPL.team.stats)
names(EPL.team.stats) <- c("Team_Name", "Team", "Points","Goal_Differential", "Mean_Goals_Scored", "Mean_Goals_Conceded")
```

### Points
```{r point_viz, echo=FALSE}
# Points
ggplot(EPL.team.stats, aes(x=Team_Name,y=Points)) + 
  geom_col() + ggtitle("Total Points Scored per Team in last past 20 Seasons (1998-99 through 2017-18)") +
  theme(plot.title=element_text(size=11)) +
  labs(x="Teams", y = "Points") +
  theme(axis.text.x=element_text(angle = 90, size = 8, vjust=-.05))
```

### Goal Differential
```{r GD_viz, echo=FALSE}
# Goal Differential
ggplot(EPL.team.stats, aes(x=Team_Name,y=Goal_Differential)) +
  geom_col() + ggtitle("Total Goal Differential Scored per Team in last past 20 Seasons (1998-99 through 2017-18)") +
  theme(plot.title=element_text(size=11)) +
  labs(x="Teams", y = "Goal Differential") +
  theme(axis.text.x=element_text(angle = 90, size = 8, vjust=-.05))
```

### Mean Goals Scored
```{r MGS_viz, echo=FALSE}
# Mean Goals Scored
ggplot(EPL.team.stats, aes(x=Team_Name,y=Mean_Goals_Scored)) +
  geom_col() + ggtitle("Mean Match Goals Scored per Team in last past 20 Seasons (1998-99 through 2017-18)") +
  theme(plot.title=element_text(size=11)) +
  labs(x="Teams", y = "Mean Goals Scored") +
  theme(axis.text.x=element_text(angle = 90, size = 8, vjust=-.05))
```

### Mean Goals Conceded
```{r MGC_viz, echo=FALSE}
# Mean Goals Conceded
ggplot(EPL.team.stats, aes(x=Team_Name,y=Mean_Goals_Conceded)) +
  geom_col() + ggtitle("Mean Match Goals Scored per Team in last past 20 Seasons (1998-99 through 2017-18)") +
  theme(plot.title=element_text(size=11)) +
  labs(x="Teams", y = "Mean Goals Conceded") +
  theme(axis.text.x=element_text(angle = 90, size = 8, vjust=-.05))
```

## Exploratory
## Markov Chain Simulation
### Southampton vs. Arsenal

```{r Markov Chain Simulation, echo=FALSE, include=FALSE}
#1 Field Items
homeTeam = 'Southampton'
awayTeam = 'Arsenal'
outputHomeTeam = c()
outputAwayTeam = c()
numberOfTimesToRunSim = 10
randomNumber = NULL
countGoalsScoredByHomeTeam = 0
countGoalsScoredByAwayTeam = 0

HG = mean(EPL$FTHG)
AG = mean(EPL$FTAG)
TG = (mean(EPL$FTHG) + mean(EPL$FTAG))/2 

GSBH = EPL$FTHG
GSBA = EPL$FTAG
GCBA = EPL$FTHG
GCBH = EPL$FTAG


EPL$HomeTeamXG <- (GSBH / TG) * (GCBA / TG) * (HG / TG) * TG
EPL$AwayTeamXG <- (GSBA / TG) * (GCBH / TG) * (AG / TG) * TG

#Simulation Loop 
for (x in 1:numberOfTimesToRunSim) {
        
        for (i in 1:length(EPL$HomeTeam)) {
                
                if(EPL$HomeTeam[i]==homeTeam){
                        
                        randomNumber = sample(1:100,1)
                        
                        if (randomNumber<=EPL$HomeTeamXG[i]*100) {
                                print("Southampton Scored")
                                countGoalsScoredByHomeTeam = countGoalsScoredByHomeTeam +1
                        }
                        }else if (EPL$AwayTeam[i]==awayTeam){
                        
                        randomNumber = sample(1:100,1)
                        
                        if(randomNumber<=EPL$AwayTeamXG[i]*100) {
                                print("Arsenal Scored")
                                countGoalsScoredByAwayTeam = countGoalsScoredByAwayTeam + 1
                }
                }
                }
}


#6
#Save Outputs

outputHomeTeam = c(outputHomeTeam, countGoalsScoredByHomeTeam)
outputAwayTeam = c(outputAwayTeam, countGoalsScoredByAwayTeam)

awayTeamGamesWon <- 0
homeTeamGamesWon <- 0

#Loop Wins for Away and Home
for (y in 1:length(outputHomeTeam)){
        
        if(outputHomeTeam[y]>outputAwayTeam[y]){
                
                homeTeamGamesWon = homeTeamGamesWon+1
                
        }else if (outputHomeTeam[y]<outputAwayTeam[y]){
                
                awayTeamGamesWon = awayTeamGamesWon+1
        }
        
}

HTGW = homeTeamGamesWon
ATGW = awayTeamGamesWon
TTGW = awayTeamGamesWon + homeTeamGamesWon

percGamesWonByHomeTeam = HTGW/TTGW
percGamesWonByAwayTeam = ATGW/TTGW

# Print Outcome
matchTeams = c(homeTeam, awayTeam)
matchOutcome = c(percGamesWonByHomeTeam, percGamesWonByAwayTeam)
print(matchOutcome)
```

```{r Markov Chain Contd}
# Bar plot
yMax = max(matchOutcome)+.1

barPlotTitle = "EPL Simulation Between Two Teams"

bp = barplot(matchOutcome, ylim = c(0.0, yMax), col="blue", main=barPlotTitle,ylab ="Win Probability", xlab ="Teams",
             names.arg=c(matchTeams), cex.names=0.9,)
text(bp, 0, round(matchOutcome, 2),cex=1,pos=3)

```

### What We Did and Why

We decided to take our efforts back to a memorable place and simulate EPL data. Based on data we pulled from http://www.football-data.co.uk/englandm.php, we gathered and cleaned EPL data from 1998-2018. The object was to project a potential season winner using historical data. The potential winner is represented using probabilites of position placement. The Poisson distribution was a logical method based on the randomness caused by the rpois function. 

In addition, we explored Markov Chain Simulation, specifically the similation between two teams. In our example we use Southampton and Arsenal to simulate a  number of iterations using a sample size of 100. The output is fairly interesting as you can see simulated scoring as the loop progresses. 

### What We Found

We found that based on the data we pulled starting in 1998, Manchester United was projected to place in the first postion and win the projected season. We created code to capture teams based on probability of placement. In later examples, Arsenal was used and presented a 28% probability of finishing in 2nd place and a 72% probability of finishing in 3rd place.

The Markov Chain Simulation between Southampton and Arsenal resulted in 100% chance of Arsenal winning, given a sample size of 100 and 10 simulations. 

### Why It Matters

This would be a helpful tool with gambling if we were to bring in betting odds and manage to indentify trends through simulations. Overall this was an exciting opportunity for us to see how simulations work in real world settings. Sports analytics is a growing industry as analytics as a whole evolves.


Statistical Note: For this example, independent Poisson distributions with means ExpHG and ExpAG have been used to simulate results from. 

For each iteration, the remaining season results are simulated, a total points and Goal difference table is created.
We calculate ranks based on points and goal differential.


