---
title: "EPL Simulation"
author: "Jack Letcher, Treva Winlock, Andrew Brill"
date: "March 12, 2018"
output: html_document
---


# English Premier League 1993-2018
## Source: http://www.football-data.co.uk/englandm.php

The Poisson distribution applies when: (1) the event is something that can be counted in whole numbers; (2) occurrences are independent, so that one occurrence neither diminishes nor increases the chance of another; (3) the average frequency of occurrence for the time period in question is known; and (4) it is possible to count how many events have occurred, such as the number of times a team wins or loses during a given season. The Poisson Distribution gives the expected frequency profile for events. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
```

## Temporary Home/Away Tables
```{r temp tables}
# Read in local .csv file 
EPL <- read.csv("EPL.csv")

# Format "Date" vector
EPL$Date <- as.Date(dmy(EPL$Date))
EPL$Date <- as.Date(EPL$Date, "%Y/%m/%d")

# Create Match and Team variables
match_table <- paste(EPL$HomeTeam, EPL$AwayTeam, sep = " - ")
team_table <- unique(c(EPL$HomeTeam, EPL$AwayTeam)) %>% sort

# Create home table (Games Played, Pts., Goals Scored, Goals Conceded)
home_table <- EPL %>% 
          group_by(HomeTeam) %>%
          summarise(P = length(FTR),
                    Pts = sum((FTHG > FTAG) * 3 + (FTHG == FTAG) * 1),
                    GS = sum(FTHG),
                    GC = sum(FTAG)) %>%
          ungroup()

# Used for a lookup table 
home_table <- cbind(home_table, Team_Num = 1:47)

# Create away table (Games Played, Pts., Goals Scored, Goals Conceded)
away_table <- EPL %>% 
          group_by(AwayTeam) %>%
          summarise(P = length(FTR),
                    Pts = sum((FTHG < FTAG) * 3 + (FTHG == FTAG) * 1),
                    GS = sum(FTAG),
                    GC = sum(FTHG)) %>%
          ungroup()

# Used for a lookup table
away_table <- cbind(away_table, Team_Num = 1:47)

# Create consolidated table with Total Goals Scored/Conceded
EPL.team.stats <- data.frame(Team = team_table,
                            Points = home_table$Pts + away_table$Pts,
                            GD = (home_table$GS + away_table$GS) - (home_table$GC + away_table$GC),
                            TGS = (home_table$GS + away_table$GS)/(home_table$P + away_table$P),
                            TGC = (home_table$GC + away_table$GC)/(home_table$P + away_table$P), stringsAsFactors = FALSE)
```

### Visualizations
```{r presultssure, echo=FALSE}
# Visualize home and away tables
plot(home_table)
plot(away_table)
```

## Matches To Be Played
```{r TBP}
# Create new EPL table that normalizes Goals Scored and Goals Conceded
# Create Expected Home Goals and Expected Away Goals
EPL.new <- expand.grid(HomeTeam = team_table, AwayTeam = team_table, stringsAsFactors = FALSE) %>%
            filter(HomeTeam != AwayTeam) %>%
            mutate(Match = paste(HomeTeam, AwayTeam, sep = " - ")) %>%
            filter(!(Match %in% match_table)) %>%
            select(-Match) %>%
            mutate(HG = mean(EPL$FTHG),
                   AG = mean(EPL$FTAG),
                   TG = (mean(EPL$FTHG) + mean(EPL$FTAG))/2) %>%
            right_join(subset(EPL.team.stats, select = -c(Points, GD)),  by = c("HomeTeam" = "Team")) %>%
            right_join(subset(EPL.team.stats, select = -c(Points, GD)), by = c("AwayTeam" = "Team")) %>%
            setNames(c("HomeTeam", "AwayTeam", "HG", "AG", "TG", 
                       "GSBH", "GCBH", "GSBA", "GCBA")) %>%
            mutate(ExpHG = (GSBH / TG) * (GCBA / TG) * (HG / TG) * TG,
                   ExpAG = (GSBA / TG) * (GCBH / TG) * (AG / TG) * TG) %>%
            ungroup()
```

## Season Simulation
```{r simulation}
# Number of games simulated
iSimulation <- 500
n <- length(team_table)

# Create new shell for simulation data
EPL.all <- data.frame(Team = rep(team_table, iSimulation),
                     SimNo = rep(1:iSimulation, each = n),
                     Pts = rep(NA, n * iSimulation),
                     GD = rep(NA, n * iSimulation),
                     Rank = rep(NA, n * iSimulation))

# Progress Bar for simulation
win_bar <- winProgressBar(title = "Churning Through Your Simulation", 
                     label = "Simulating ... 0% done", 
                     min = 0, 
                     max = iSimulation, 
                     initial = 0)

# For loop that simulates through preset number of loops
for (i in 1:iSimulation){

# Random Generation for Poisson Distribution -- rpois
# Lambda representing vector of non-negative means         
  temp_table <- EPL.new %>% 
              mutate(x1 = rpois(nrow(EPL.new), lambda = EPL.new$ExpHG), 
                     x2 = rpois(nrow(EPL.new), lambda = EPL.new$ExpAG), 
                     HPts = 3 * (x1 > x2) + 1 * (x1 == x2),
                     APts = 3 * (x1 < x2) + 1 * (x1 == x2))
# Results table   
  results <- EPL.team.stats %>% select(Points, GD) + 
         temp_table %>% group_by(HomeTeam) %>% summarise(Pts = sum(HPts),
                                                  GD = sum(x1) - sum(x2)) %>% select(Pts, GD) + 
         temp_table %>% group_by(AwayTeam) %>% summarise(Pts = sum(APts),
                                                  GD = sum(x2) - sum(x1)) %>% select(Pts, GD) 

  EPL.all[(n*(i-1) + 1):(n*i), c("Pts", "GD")] <- results
  
  results$PGD <- results$Points + (results$GD - min(results$GD) + 1) / max((results$GD - min(results$GD) + 1) + 1)
  EPL.all[(n*(i-1) + 1):(n*i), c("Rank")] <- rank(-results$PGD, ties.method = "random")  

# Progress Bar for simulation results  
  info <- sprintf("Simulating ... %d%% complete", round((i/iSimulation)*100))
  setWinProgressBar(win_bar, i, label = info)  
}
close(win_bar)
```


```{r teamname_addition, echo=FALSE}
# Create new table to house lookup values for teamnames
new <- EPL.all
new[] <- home_table$HomeTeam[match(unlist(EPL.all), home_table$Team_Num)]

# Combine tables to clean up numeric teams
EPL.all.new <- cbind(Team_Name = new[1], EPL.all)
EPL.all.new <- EPL.all.new[-c(2)]
```

## Winner
```{r winner model}
# Output for Position, also provides probability
EPL.all.new %>% filter(Rank == 1) %>% select(Team) %>% table/iSimulation
```

## Arsenal Projected Finish
```{r projected }
# Output for Team, also provides probability
EPL.all.new %>% filter(Team == "Arsenal") %>% select(Rank) %>% table/iSimulation
```

## Arsenal Point Distribution
```{r point distribution}
# Output for Point Distribution, also provides probability
EPL.all.new %>% filter(Team == "Arsenal") %>% select(Pts) %>% table/iSimulation
```
## Table of Probabilities per Position
```{r Probabilities}
# Full table results by Team and probabilties of finish place
table(EPL.all.new$Team, EPL.all.new$Rank)/iSimulation
```

Statistical Note: For this example, independent Poisson distributions with means ExpHG and ExpAG have been used to simulate resultsults from. 

For each iteration, the remaining season results are simulated, a total points and Goal difference table is created.
We calculate ranks based on points and goal differential.


